---
title: "Conclusion"
format:
  html: 
    toc: false
    link-external-icon: true
    link-external-newwindow: true
    link-external-filter: ^(?:http:|https:)\/\/www\.openstatprog\.com
---

```{r include = FALSE}
source(here::here("R", "buttons.R"))
```

<!--
## Hand-In Assignment

::: practice

## Vectors & Matrices Assignment

Complete each step below in your own R session, then submit your work (see "What to Hand In" below).

1. **Generate.** Create a vector called `temps` containing the integer sequence from 60 to 90, in steps of 5 (i.e.,
   60, 65, 70, ..., 90).
1. **Operate & explain (recycling).** Create a second vector called `adjust` containing the two values `-2` and `3`.
   Compute `temps + adjust` and store the result as `adjusted_temps`. In a comment, explain in your own words *why*
   this produces a length-7 result from a length-7 vector and a length-2 vector, and whether R issued a warning (and
   why or why not).
1. **Select & modify.** Using indexing (not retyping the whole vector), replace every value in `adjusted_temps` that
   is greater than 80 with the value `80` (i.e., "cap" the temperatures at 80).
1. **Matrices.** Create a 7×2 matrix called `temp_matrix` where the first column is `temps` and the second column is
   `adjusted_temps`. (Hint: `cbind()` can combine vectors into the columns of a matrix, or you can build it with
   `matrix()` directly — either is fine.)
1. **Conceptual explanation.** In a comment, explain in your own words how `temp_matrix` is related to the two
   vectors you built it from — specifically, what R is doing "under the hood" that makes a matrix possible.

### What to Hand In

Save all of your commands, in the order you ran them, in a single plain-text or R script file (e.g.
`yourname_vectors-matrices.R`), including both required comments. 

A teacher reviewing your submission should be able to see: correct vector generation, a working (and correctly explained) recycling operation, correct conditional selection/modification, a correctly constructed matrix, and a clear explanation of the vector–matrix relationship in your own words.

:::
-->

Congratulations! You've completed the **Vectors & Matrices** module.

## Learning Goals

Now that you've completed this module, you should be able to:

### Vectors

1. Describe the characteristics of atomic vectors.
1. Describe the six atomic vector modes (logical, integer, double, character, complex, raw).
1. Generate new atomic vectors using various functions.
1. Perform element-wise arithmetic and logical operations on atomic vectors.
1. Select and modify elements in a vector.
1. Explain "recycling", and describe how it works when manipulating vectors.

### Matrices

1. Describe the characteristics of an R matrix. 
1. Explain the relationship between matrices and vectors. 
1. Create new matrices in R.
1. Perform element-wise operations on matrices. 
1. Select and modify the elements of a matrix.
1. Explain how recycling works with matrix operations. 

## Additional Reading

For more information on the topics we've covered in this module, the following chapter from Hadley Wickham's *R for
Data Science* is an excellent resource.

- [Vectors](https://r4ds.had.co.nz/vectors.html) in the [First Edition](https://r4ds.had.co.nz/index.html).

### Advanced Reading

For a much more technical deep-dive into the topics we've covered in this module, check out the following chapter
from Hadley Wickham's [Advanced R](https://adv-r.hadley.nz/index.html).

- [Vectors](https://adv-r.hadley.nz/vectors-chap.html)

## Next Steps 

Now that you have a solid grasp of vectors and matrices, you're ready to learn about two very useful data structures:
lists and data frames. These data types are central to any real-world R project because they can store let you store
mixed-type, tabular data and arbitrary collections of R objects.

```{r}
#| echo: false
#| results: asis

next_module_button("https://www.openstatprog.com/lists-data-frames/index.html")
```
