---
title: "Vectors & Matrices"
page-navigation: false
format:
  html: 
    toc: false
---

```{r, echo = FALSE, include = FALSE}
source(here::here("R", "buttons.R"))
```

::: {.callout-tip}

## ⏱ Estimated Time to Complete

**About 90–120 minutes**, including practice exercises and both knowledge quizzes.

:::

In this module, we'll cover two of R's most fundamental data structures: **vectors** and **matrices**. These
structures form the foundation of nearly every data manipulation and data analysis workflow in R and everything
else you'll encounter later in other modules (lists, data frames, factors) is either built out of vectors or behaves
according to rules you first learn here. Understanding vectors and matrices thoroughly now will make everything that
follows easier.

Content is split into two tutorials:

1. [Vectors](/tutorials/vectors/index.qmd)
1. [Matrices](/tutorials/matrices/index.qmd)

## Learning Goals

After completing this module, you will be able to:

### Vectors

1. Describe the characteristics of atomic vectors, and explain their fundamental role in the R data system.
1. Describe the six atomic vector modes (logical, integer, double, character, complex, raw).
1. Generate new atomic vectors using various functions.
1. Perform element-wise arithmetic and logical operations on atomic vectors, and explain how recycling works.
1. Select and modify elements in a vector.

### Matrices

1. Describe the characteristics of an R matrix. 
1. Explain the relationship between matrices and vectors. 
1. Create new matrices in R.
1. Perform element-wise operations on matrices. 
1. Explain how recycling works with matrix operations. 
1. Select and modify the elements of a matrix.

Click the button below to get started with the first tutorial.

```{r}
#| echo: false
#| results: asis
begin_button("/tutorials/vectors/index.html")
```
