---
title: "Matrix Operations"
---

```{r, echo = FALSE, include = FALSE}
set.seed(235711)
```

## Size

The `length()` function works with matrices, but the results may not be that interesting.

```{r}
(m1 <- matrix(1:9, 3, 3))
length(m1)
```

**Do:** Explain what the value returned by `length(m1)` represents.

::: explanation

When applied to matrices, `length()` returns the number of cells in the matrix.

:::

To get more useful information about the dimensions of the matrix, we use `dim()`, `nrow()`, or `ncol()`.

```{r}
# Number or rows and columns
dim(m1)

# Number of rows
nrow(m1)

# Number or columns
ncol(m1)
```

## Arithmetic

As with vectors, default arithmetic with R matrices works *element-wise*. R performs the requested operation on each
pair of corresponding entries in the two matrices. 

```{r}
(m2 <- matrix(sample(-5:5, 9, TRUE), 3, 3))

m1 + m2
m2 - m1
m1 / m2
m1 * m2
```

::: practice

Use the three matrices defined below to answer this question.

```{r}
a <- matrix(2, 2, 2)
a

b <- matrix(1:4, 2, 2)
b

c <- matrix(6, 2, 2)
c
```

:::: {.panel-tabset}

## {{< var tab_title.question >}}

Given the above definitions of `a`, `b`, and `c`, **predict** what the following code will print.

```{r}
#| eval: false

b + c / a
```

Use the interactive editor to check your prediction.

## {{< var tab_title.hint >}}

R will apply the usual order-of-operations when doing element-wise arithmetic with matrices.

## {{< var tab_title.editor >}}

```{webr-r}
#| context: setup

a <- matrix(2, 2, 2)
a

b <- matrix(1:4, 2, 2)
b

c <- matrix(6, 2, 2)
c
```

```{webr-r}
```

## {{< var tab_title.solution >}}

We get the following matrix.

```{r}
b + c / a
```

R first divides each element in `c` by the corresponding element in `a`.

```{r}
(d <- c / a)
```

Then, the result of this division is added to `b`.

```{r}
b + d
```

::::

:::

::: going-further

## Matrix Algebra

If you're familiar with matrix algebra and/or have some affinity for programming languages that overload their
operators, the matrix arithmetic described above may seem very strange. Rather than overloading the standard operators,
R defines special functions for matrix algebraic operations. For example:

- `%*%`: Multiplication
- `t()`: Transposition
- `solve()`: Inversion

```{r}
# Multiply m1 and m2 using true matrix multiplication
m1 %*% m2

# Transpose m1
t(m1)

# Invert m2
solve(m2)
```

:::

::: practice

1. Create a $5 \times 3$ numeric matrix called `myMat` wherein each column is equal to the vector `1:5`.
1. Multiply each entry in `myMat` by $\pi$ (i.e., the mathematical constant, "pi").

:::: {.panel-tabset}

## {{< var tab_title.editor >}}

```{webr-r}
```

## {{< var tab_title.hint >}}

The built-in R object `pi` contains the value of $\pi$.

## {{< var tab_title.solution >}}

```{r}
(myMat <- matrix(1:5, 5, 3))

myMat * pi
```

::::

:::

::: knowledge-check


:::
