---
title: "Manipulating Matrices"
---

::: {.callout-note}
Selecting and modifying matrix elements is a hands-on, procedural skill, directly extending
what you already practiced with vectors.
:::

## Accessing Matrix Elements

As with vectors, we can access and assign values within a matrix using the square brackets operator, `[]`, but with one key difference: matrices require two indices, one to define the row selection and one to define the column selection.

```{r}
(m1 <- matrix(1:12, 3, 4))

# Select the first two rows and the first three columns, then print the selected
# elements as a matrix
m1[1:2, 1:3]

# Select the first and third rows and second and third column, then save the
# selection as a new matrix called `m2`
(m2 <- m1[c(1, 3), 2:3])
```

## Modifying Matrix Elements

Editing the value of matrix elements works the same way it did for vectors. We , use the `[]` operator to select the values we want to overwrite, and assign new values for the selected elements using the ordinary assignment operator,
`<-`.

```{r}
(m3 <- matrix(1, 3, 3))

# Replace the element in the first row and second column with 33
m3[1, 2] <- 33
m3

# Replace all elements in the range covered by the second and third rows and the
# first and third columns with 44 
m3[2:3, c(1, 3)] <- 44
m3
```

::: practice

## Practice: Predict, Then Run

Given `m1 <- matrix(1:12, 3, 4)` (already defined above), predict what `m1[, 2]` will return before running it. Pay attention to whether the result is a matrix or a plain vector.

:::: {.panel-tabset}
## {{< var tab_title.question >}}

```{r, eval = FALSE}
m1[, 2]
```

## {{< var tab_title.solution >}}

```{r}
m1[, 2]
```

Leaving the row index blank (`m1[, 2]`) tells R to select *all* rows, restricted to column 2. Notice that R returns a plain vector here, not a one-column matrix — by default, R "drops" a dimension when a selection only has one column (or one row). We'll see this same behavior again with data frames in the next module.
::::

:::

## Quick Check

- What's the difference between `m1[1, ]` and `m1[, 1]`?
- Why does selecting a single row or column from a matrix return a plain vector instead of a matrix?

```{webr-r}
```

::: {.callout-tip}

## Ready to Continue?

You've now completed the *Vectors & Matrices* module. You're ready to move on if you can:

- Explain what makes a matrix "just a vector with a `dim` attribute."
- Create a matrix with `matrix()`, and predict whether it fills by row or by column.
- Apply arithmetic to matrices and predict when recycling will happen silently vs. with a warning.
- Select and modify matrix elements using row and column indices.

If any of these still feel shaky, revisit the relevant page above, or try the
[Knowledge Quiz](/quizzes/matrices/index.qmd) to pinpoint exactly where to focus. Once you're ready, head to the
[Conclusion](/conclusion.qmd) page to find this module's hand-in assignment.

:::
