---
title: "Creating Matrices"
---

::: {.callout-note}

Creating matrices is procedural, but understanding *why* a matrix is "just a vector with a `dim` attribute" is
conceptual, and that concept is the single most important idea on this page.

:::

The most direct way to create a new matrix is the `matrix()` function. When we create a matrix, we need to provide some
data for the matrix to hold, and we need to tell R how many rows and columns the matrix should contain.

```{r}
(m1 <- matrix(data = 1, nrow = 3, ncol = 3))
```

If we inspect the object, we’ll see it now has an *attribute*, `dim`. The attribute `dim` is a two element vector, in
which the first element shows the number of rows and the second element the number of columns.

```{r}
# Matrices must have at least a 'dim' attribute
attributes(m1)
```

Vectors, on the other hand, don't have attributes.

```{r}
# Create a numeric vector for comparison
y1 <- c(1, 2, 3)

# Basic vectors don't have attributes
attributes(y1)
```

## The Matrix/Vector Relation

I'm not being facetious when I say that a matrix is just a vector with a `dim` attribute. In fact, we can convert a
vector to a matrix simply by adding a `dim` attribute to the vector.

### Vector $\rightarrow$ Matrix

**See:** Notice that `y1` is an ordinary numeric vector.

```{r}
class(y1)
is.vector(y1)
```

Unsurprisingly, `y1` is not a matrix.

```{r}
is.matrix(y1)
```

But we can change all that by adding a `dim` attribute to `y1`.

```{r}
attr(y1, "dim") <- c(3, 1)
y1
```

Now, `y1` looks like a $3 \times 1$ matrix when printed. Indeed, as far as R is concerned, `y1` is now a matrix and no
longer a vector.

```{r}
class(y1)
is.matrix(y1)
is.vector(y1)
```

### Matrix $\rightarrow$ Vector

Similarly, we can convert a matrix to a vector by removing the `dim` attribute from the matrix.

**See:** Notice that `m1` is a matrix and not a vector.

```{r}
class(m1)
is.matrix(m1)
is.vector(m1)
```

As above, we'll transmute `m1` by manipulating its `dim` attribute. Specifically, we will remove the `dim` attribute
entirely.

```{r}
# Setting the 'dim' attribute of m1 to NULL effectively removes that attribute
attr(m1, "dim") <- NULL
m1
```

Now, `m1` prints like a length-nine vector, and R agrees: `m1` is now a vector, not a matrix.

```{r}
class(m1)
is.vector(m1)
is.matrix(m1)
```

## Element Ordering

By default, R fills matrices column-wise (i.e., using *column-major order*): the first column is filled from top to
bottom, then the second column is filled top to bottom, then the third column, and so on.

```{r}
matrix(1:9, 3, 3)
```

If we want to fill the matrix row-by-row instead (i.e., using *row-major order*), we can use the `byrow = TRUE` argument.

```{r}
matrix(1:9, 3, 3, byrow = TRUE)
```

::: practice

**Predict** what the matrix generated by `matrix(1:6, nrow = 2, ncol = 3)` will look like *before* running any code.

- Use the interactive editor to check your prediction.

:::: {.panel-tabset}

## {{< var tab_title.editor >}}

```{webr-r}
matrix(1:6, nrow = 2, ncol = 3)
```

## {{< var tab_title.solution >}}

```{r}
matrix(1:6, nrow = 2, ncol = 3)
```

Since R fills matrices column-by-column:

- `1` and `2` fill the first column
- `3` and `4` fill the second column
- `5` and `6` fill the third column

::::

:::

::: knowledge-check

- In one sentence, explain what makes an R matrix different from an R vector.
- If you remove a matrix's `dim` attribute, what do you get back?
- If you don't specify any special arguments, does R fill a matrix by row or by column?

```{webr-r}
```

:::
