---
title: "Generating Vectors"
---

::: {.callout-note}

This page is hands-on: you'll write code that generates vectors in several different ways.

:::

There are many different ways to generate vectors in R. Some involve combining explicitly specified elements, while
others use algorithmic rules to generate values that follow some pattern. Below, we'll cover some of the most common
ways to generate vectors.

## Concatenation

The *concatenation* function, `c()`, constructs a vector by joining its arguments into a single vector.

```{r}
(y1 <- c(1, 2, 3))
(y2 <- c(TRUE, FALSE, TRUE, TRUE))
(y3 <- c("bob", "suzy", "danny"))
```

::: {.column-margin}

:::: {.callout-note}

The outer parentheses in these examples are just a handy piece of syntactic sugar. They allow us to assign a value
(e.g., `c(1, 2, 3)`) to an object (e.g., `y1`) and print that object in a single command. You don't need to use these
extra parentheses if you don't want to print the value of object your creating.

::::

:::

**Do:** Run the following code to create an integer vector using the `c()` function.

```{webr-r}
c(2, 3, 5, 7, 11)
```

**See:** The command above generates an integer vector containing the first five positive, prime numbers. Notice that
the vector is printed immediately because we did not assign the output of `c()` to any object.

**Do:** Modify the above code to save the vector as an object called `y` and printing the contents of `y`, as two
separate steps.

::: explanation

```{r}
y <- c(2, 3, 5, 7, 11)
y
```

:::

## Generating Sequences

The colon operator, `:`, creates a sequence of regularly spaced values between two endpoints. Each element in the
resulting vector is separated from it's neighbors by $\pm$ 1.

```{r}
1:5
```

You need to be careful when using negative numbers with the `:` operator. Notice how the placement of the negative
values in the following two commands affect the resulting sequences.

```{r}
-3:2
1:-2
```

::: practice

:::: {.panel-tabset}

The `:` operator is intended for creating sequences of whole numbers, but it will also work with decimal-valued
arguments, though the results might not be what you'd expect.

## {{< var tab_title.question >}}

Without running the code, see if you can generalize what you know about the `:` operator to answer the following
questions.

- What vector will the following code produce?
- What rule is `:` using to generate this vector?

```{r, eval = FALSE}
1.2:5.3
```

## {{< var tab_title.solution >}}

The resulting vector looks like this:

```{r, echo = FALSE}
1.2:5.3
```

This vector is generated by assigning 1.2 as the first element and then adding 1 to define the next element. Each
subsequent element is defined by adding one to the last element until the resulting value would exceed the upper bound
we requested (i.e., 5.3). So, the last element in the vector is 5.2, since that's the largest value that can be
generated in unit steps from 1.2 without exceeding 5.3.

::::

:::

To generate regular sequences with non-unit steps between the elements, we can use `seq()`. The `seq()` function allows
you to specify the step size or the total number of elements in the resulting vector with the `by` or `length.out`
arguments, respectively.

```{r}
# Create a numeric vector starting at 0, ending at 1, with elements
# increasing by steps of 0.25
seq(from = 0, to = 1, by = 0.25)

# The same as above but omitting the argument names
seq(0, 1, 0.25)

# Create a numeric vector of 10 evenly spaced values starting at -1 and
# ending at 1
seq(-1, 1, length.out = 10)
```

::: practice

:::: {.panel-tabset}

Use the `seq()` function to complete these tasks.

## {{< var tab_title.question >}}

1. Generate a vector that begins with 7, ends with 42, and contains 8 evenly spaced elements.
1. Generate a vector that contains every odd integer between 3 and 17 (inclusive).

## {{< var tab_title.editor >}}

```{webr-r}
```

## {{< var tab_title.solution >}}

The `length.out` argument provides the most direct solution for the first problem.

```{r}
seq(7, 42, length.out = 8)
```

The second problem is easiest to solve with the `by` argument.

- Using a step size of 2 includes only odd values.

```{r}
seq(3, 17, 2)
```

::::

:::

## Repeating Values

The `rep()` function creates vectors by repeating it's first argument as many times as requested.

```{r}
rep(33, times = 4)
rep(1:3, 3)
rep(c("foo", "bar"), 4)
```

Use the `each` argument to repeat each element of the first argument `each` number of times.

```{r}
rep(1:3, each = 3)
rep(c(TRUE, FALSE), each = 2)
```

We can use `rep()` to generate non-numeric vectors, too.

**Predict:** What will the two following commands print?

```{webr-r}
rep(c("foo", "bar"), 4)
rep(c(TRUE, FALSE), each = 2)
```

::: explanation

- The first line prints a length-eight character vector wherein the set {`"foo"`, `"bar"`} is repeated four times.
- The second line prints a length-four logical vector wherein `TRUE` is first repeated twice, and then `FALSE` is
repeated twice.

:::

::: practice

Use the `rep()` function to generate the following two vectors.

```{r}
#| echo: false

rep(c(2, 3, 5), 2)
rep(c("alice", "bob"), each = 3)
```

:::: {.panel-tabset}

## {{< var tab_title.editor >}}

```{webr-r}
```

## {{< var tab_title.solution >}}

Using the `times` argument is best for the first vector.

```{r}
rep(c(2, 3, 5), 2)
```

For the second vector, it's easier to use the `each` argument.

```{r}
rep(c("alice", "bob"), each = 3)
```

::::

:::

::: knowledge-check

- Which function would you use to generate the sequence `2, 4, 6, 8, 10`
   - `c()`
   - `:`
   - `seq()`

   Is there more than one right answer?

- In the `rep()` function, what's the difference between the `each` and `times` arguments?

```{webr-r}
```

:::
