---
title: "The Role of Vectors"
---

::: going-further

The information on this page is largely option extra detail about the role that vectors play in the overall design
philosophy of R. Understanding these concepts will help you make sense of the more complex types of objects that you'll
encounter in later modules. These ideas can also be helpful when you're trying to figure out why R is doing some
confusing nonsense that you didn't expect. All that being said, however, you do not need to master these concepts to
progress to the next tutorial.

:::

Before we proceed to more complex types of data objects, let's take a moment to consider the unique role that vectors
play in R's design philosophy. In particular, we should explore what makes vectors "atomic" data types.

## How are Vectors Atomic?

R vectors are "atomic" in two ways.

1. Vectors are irreducible.
1. Vectors are homogeneous.

### Irreducibility

As we briefly mentioned on the first page of this tutorial, vectors are the simplest data type in R. So, there is no
such thing as a true scalar in R. By the standards of most programming languages, that's very strange. In R, any single
value with one of the six atomic modes is a length-one vector. So, as far as R is concerned the following expressions do
not represent a number, a boolean value, and a character string.

```{r}
#| eval: false

3.14
TRUE
"foo bar"
```

R treats each of the above as a different flavor of length-one vector.

#### Numbers

R views one number as a length-one numeric vector.

```{r}
typeof(3.14)
length(3.14)
is.vector(3.14)
```

#### Boolean Values

R views one `TRUE` or `FALSE` as a length-one logical vector.

```{r}
typeof(TRUE)
length(TRUE)
is.vector(TRUE)
```

#### Character Strings

R views a single quoted piece of text as a length-one character vector.

```{r}
typeof("foo bar")
length("foo bar")
is.vector("foo bar")
```

For most of your day-to-day R programming and data analysis, the technical distinction between a scalar and a length-one
vector won't matter. Sometimes, however, remember that vectors act as R's most elemental building blocks can help you
understand the behavior of more complex objects (e.g., lists and data frames).

### Homogeneity

We briefly addressed the homogeneity of vectors in the [Atomic Modes](/tutorials/vectors/atomic_modes.qmd) section, but
the concept is worth revisiting. When we say that vectors are homogeneous, we mean a vector can only contain one type of
data. Hopefully, that restriction seems obvious when you consider that vectors are the simplest possible type of object
in R. If vectors could store different types of data, we'd need some simpler type of object to represent the pieces of
heterogeneous data we want to store in our vector. Ergo, since there's nothing simpler than a vector, vectors must
contain only one type of data.

#### Type Casting

Considering the draconian restrictions on vector types, it may surprise you to learn that the following code is
perfectly legal and won't cause any errors. We're attempting to concatenating a length-one numeric vector, a length-one
logical vector, and a length-one character vector into a single length-three vector. Why does that work?

```{r}
c(1, FALSE, "foo")
```

When you try to combine heterogeneous data into a single vector, R will automatically convert the types of some elements
to produce a legal atomic vector. In the example above, the only way to resolve the type discrepancy was to quote each
element and return a character vector.

In the following example, R was able to resolve the type discrepancy with a less drastic conversion. Merely converting
`FALSE` to `0` and returning a numeric vector was enough.

```{r}
c(1, 2, FALSE)
```

Note that R will almost always honor quotes, and won't automatically convert character strings to any other type, even
when as sensible conversion may seem obvious.

```{r}
c("1", 2, FALSE)
c("1", 2, 3)
c("TRUE", FALSE, FALSE)
```

We can trigger the same automatic type casting by carelessly overwriting vector elements. For example, in the following
code, the logical value `TRUE` that we attempt to assign to the fourth slot in `x` is automatically converted to the
integer `1` to conform to the atomic mode of `x`.

```{r}
(x <- 1:5)
x[4] <- TRUE
x
```

Conversely, in the example below, when we try to assign the numeric value `1` to the second slot in the logical vector
`y`, R automatically converts `y` to a numeric vector to accommodate the new data. 

```{r}
(y <- rep(c(TRUE, FALSE), 3))
y[2] <- 1
y
```

::: knowledge-check

- What is the simplest type of data object that R recognizes?
- In terms of R's type system, how would you describe this object: `"foo"`?
- What type of vector would the following command return `c(FALSE, TRUE, "bob")`?

```{webr-r}
```

:::

::: {.callout-tip}

## Ready to Continue?

You're ready to move on to [Matrices](/tutorials/matrices/index.qmd) if you can:

- Name the six atomic vector modes, and describe what type of data each mode stores.
- Generate vectors using `c()`, `:`, `seq()`, and `rep()`.
- Explain, in your own words, what recycling is, when it happens, and why it can be useful.
- Select and modify specific elements of a vector using `[]`.

Take the [Knowledge Quiz](/quizzes/vectors/index.qmd) to check your learning.

:::
