---
title: "Vector Operations"
---

R vectors are dimensionless objects. Unlike mathematical column vectors, R vectors don't have rows or columns, but they
do have length. We can check the length of a vector using the `length()` function.

```{r}
y <- 1:4
length(y)
```

We can check the atomic mode of a vector (i.e., what type of data it represents) with the `typeof()` function.

```{r}
typeof(1:3)
typeof("foo")
typeof(FALSE)
```

**Predict:** What will the following code print?

```{webr-r}
x <- vector("complex", 3)
typeof(x)
```

::: explanation

The first line uses `vector()` to create a length-three complex vector called `x`. So, the second line prints
`"complex"`, since that's the atomic mode of `x`. 

:::

## Arithmetic

Arithmetic in R works element-wise when applied to vectors. That is, the requested operation is performed between
matched elements of the vectors.

::: {.column-margin}

:::: going-further

If you're familiar with matrix algebra or other programming languages, the way R handles arithmetic with vectors may
seem strange. For arithmetic operations, R vectors behave much like arrays would in other programming languages.

::::

:::

```{r}
(x <- rep(2, 4))

y + x
y - x
y / x
y * x
```

### Recycling

::: {.callout-important}

## Tricky Topic

Recycling is a ubiquitous, and often silent, feature of R. Recycling is often happening behind the scenes, and you won't
be explicitly notified (via a warning or error) in most cases. Take your time with this section and try to really
understand the concept of recycling before progressing.

:::

If two vectors involved in an element-wise operation have different lengths, R will *recycle* the elements of the shorter
vector until each element in the longer vector can be matched with an element from the shorter vector.


In the following code chunk, R is using recycling to calculate the elements of `w`. Can you see how?

**Do:** Explain how the recycling algorithm works in this example.

```{r}
z <- c(1, 2)
w <- y - z
w
```

::: explanation

In this example, `y` has four elements, but `z` has only two. So, R re-uses the two elements in `z` to define matches
for all four elements in `y` and specify the four differences needed to define the elements of `w`.

1. $w_1 = y_1 - z_1$
1. $w_2 = y_2 - z_2$ $\leftarrow$ After calculating the second difference, we've run out of elements in `z`.
1. $w_3 = y_3 - z_1$ $\leftarrow$ Beginning with the third difference, we start recycling the elements of `z`.
1. $w_4 = y_4 - z_2$

:::

::: {.callout-important}

Notice how the example above runs without any explicit indication of that recycling has been used. If the length of the
longer vector is divisible by the length of the shorter vector, R will recycle silently (i.e., without any messages,
warnings, or errors). So, you need to vigilant and make sure you don't accidentally trigger recycling when you don't
want it.

:::

In cases where the length of the shorter vector doesn't evenly divide the length of the longer vector, R will still use
execute the requested operation and apply recycling, but it will also return a warning.

```{r, warning = TRUE}
a <- 1:4
b <- 1:2
c <- 1:5

# Silent recycling
a + b

# Recycling with warning
a + c
```

#### Why Recycling is a Useful Feature

At this point, if you have any predilection toward strictly typed programming languages, all this recycling business
might sound like insanity, but there's a method to the madness. Let's consider a few examples that demonstrate how 
recycling can actually be very helpful in day-to-day programmatic data analysis.

The following code looks like it should implement scalar multiplication, and outcome looks like the result of a scalar
multiplication. But R doesn't have scalars, and vector arithmetic in R doesn't follow standard linear algebraic rules,
so our intuitive reading of this expression can't be correct. 

```{r}
3 * y
```

In fact, the preceding code is actually doing element-wise multiplication of a length-one vector, `3`, and a length-four
vector, `y`, and recycling the elements of the length-one vector to resolve the length difference. In this example, 
recycling provides two benefits:

1. We get to write code that reads like the mathematical representation of the operation we're trying to implement
1. Our code produces the same result as the mathematical operation it appears to represent

The following three expressions follow the same pattern as the last example but provide more compelling evidence for the
real-world usefulness of recycling. These expressions don't represent valid operations under the conventional rules of
linear algebra, and they wouldn't be valid commands in a highly structure programming language like C. Yet, the intended
operations are intuitively obvious, and R will use recycling to execute the operations we expect.

```{r}
y - 1
y + 3
y / 3
```

## Logical Tests

Logical comparisons between vectors test some logical relation between the vectors and return a logical vector that
encodes the answer to whatever TRUE/FALSE question the logical relation tested. These comparisons follow the same rules
as vector arithmetic:

- Operators are applied element-wise
- Recycling is used to resolve length differences

```{r}
# Check which elements of y are equal to 4
y == 4

# Check which elements of y are equal to 4 or 2
y == 4 | y == 2
```

We need to be especially mindful of the vector operation rules when applying logical comparisons to vectors. When defining
logical tests, intuition can easily fail us. For example, it would be natural to read the following code as implementing
the same test as the previous expression, but the two results are clearly different. Why?

```{r}
y == c(4, 2)
```

As you've probably surmised, recycling is causing the discrepancy. We can clarify the differences between the two tests
by extending the shorter vectors to manually implement recycling. 

```{r}
y == c(4, 4, 4, 4) | y == c(2, 2, 2, 2)
y == c(4, 2, 4, 2)
```

Now, the two tests don't appear very similar, and their different outcomes aren't particularly surprising.

::: practice

## Practice

1. Create the object `myVec` by running the two lines of code below.
1. Programmatically create a logical vector that indicates which elements of `myVec` are (strictly) less than 3.

:::: {.panel-tabset}

## {{< var tab_title.question >}}

```{webr-r}
set.seed(235711)
myVec <- sample(1:5)
```

## {{< var tab_title.solution >}}

```{r}
set.seed(235711)
myVec <- sample(1:5)

y <- myVec < 3
```

::::

:::

## Quick Check

In your own words (no code needed):

- What does it mean for R to "recycle" a vector, and when does it happen?
- Why does R warn you in some recycling cases but not others?
- `y == c(4, 2)` and `y == 4 | y == 2` look like they should test the same thing, but they don't. Explain why.
